---
title: "SDSS DR12 - Fetching data using R"
author: "Bart Buelens"
date: "Thursday, February 19, 2015"
output: html_document
---

Data Release 12 (DR12) of the Sloan Digital Sky Survey is now available (early 2015). We obtain some data using SQL queries executed from within R and conduct an initial exploration of the data.

Fetching data using SQL queries
-------------------------------

First we load some libraries used in this document.

```{r, message=FALSE, cache=TRUE}
library(RCurl)
library(ggplot2)
```

We formulate a SQL query, for example the following.

```{r, cache=TRUE}
mySqlQuery = 
"SELECT objID, htmID, z, ra, dec, cx, cy, cz, class, subClass, type, survey, programname
FROM SpecPhoto
WHERE (htmid*37 & 0x000000000000FFFF) < (65)"
```

This query will ... etc + explain random number thing.

```{r, cache=TRUE}
mySqlQuery = gsub(pattern="\n",replacement=" ",x=mySqlQuery)
urlBase = "http://skyserver.sdss.org/dr12/en/tools/search/x_sql.aspx"
X = getForm(urlBase, cmd = mySqlQuery, format = "csv")
write(X,file="SDSSsample.csv")
```

We now have a data file containg the requested SDSS data. The above need only to be conducted once. Thereafter, the data can be loaded from the local hard disk, of course.

Initial data exploration
------------------------

The only aim of this short section is to ascertain that the SQL query has produced sensible data that can be used in other analyses.

Let's start clean and load the data file.

```{r, cache=TRUE}
rm(list=ls()) # clear everything in the workspace
A = read.table("SDSSsample.csv", header=TRUE, sep=",", dec=".", comment.char="#")
```

We now have data frame A containing our data.

```{r, cache=TRUE}
str(A)
xtabs(~class, A)
xtabs(~survey, A)
xtabs(~survey+class, A)
```

Plot positions on celestial sphere using the Mollweide projection. Colour code objects by their class: galaxy, quasar or star.

```{r, cache=TRUE}
ggplot() +
   geom_point(data=A, aes(x=ra, y=dec, colour=class), size=2) +
   scale_colour_hue(l=50) +
   coord_map("mollweide",xlim=c(0,360),ylim=c(-90,90)) +
   scale_y_continuous(breaks = seq(-90, 90, 30)) +
   scale_x_continuous(breaks = seq(0, 360, 60)) +
   theme(axis.title = element_blank()) +
   theme(axis.ticks = element_blank(), axis.text = element_blank()) + 
   theme(panel.border = element_blank()) + 
   theme(panel.grid.major = element_line(colour="darkgrey")) + 
   ggtitle("SDSS DR12 coverage") + 
   theme(plot.title = element_text(size = rel(2)))

ggplot() +
   geom_point(data=A, aes(x=ra, y=dec, colour=class), size=2) +
   scale_colour_hue(l=50) +
   coord_map("mollweide",xlim=c(0,360),ylim=c(-90,90)) +
   scale_y_continuous(breaks = seq(-90, 90, 30)) +
   scale_x_continuous(breaks = seq(0, 360, 60)) +
   theme(axis.title = element_blank()) +
   theme(axis.ticks = element_blank(), axis.text = element_blank()) + 
   theme(panel.border = element_blank()) + 
   theme(panel.grid.major = element_line(colour="darkgrey")) + 
   ggtitle("SDSS DR12 coverage") + 
   theme(plot.title = element_text(size = rel(1))) + 
   theme(legend.position="none") +
   facet_wrap(~class, nrow=1)
```









